\documentclass[12pt,a4paper,oneside]{report}

% ========================================
% Packages
% ========================================
% \usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[english]{babel}
\usepackage{CJKutf8}
\usepackage{xcolor}
\usepackage{csquotes}

% Page layout
\usepackage[top=25mm, bottom=20mm, left=20mm, right=20mm]{geometry}

% Graphics and figures
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}

% Mathematics
\usepackage{amsmath,amssymb,amsthm}

% Tables
\usepackage{booktabs}
\usepackage{multirow}

% References and citations
\usepackage[
    backend=biber,
    style=numeric-comp,
    sorting=none,
    maxbibnames=6,
    minbibnames=6,
    giveninits=true,
    doi=true,
    url=true,
    isbn=false
]{biblatex}
\addbibresource{references.bib}

% Line spacing
\usepackage{setspace}
\setstretch{1.3}

% Headers and footers
\usepackage{fancyhdr}
\setlength{\headheight}{14.5pt}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[R]{\thepage}
\fancyhead[L]{\leftmark}
\renewcommand{\headrulewidth}{0.4pt}

% Hyperlinks - 作業用にリンク有効
% 注: CD提出時は hyperref を無効化するか [hidelinks] オプションを使用
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
% \usepackage[hidelinks]{hyperref}

% Theorem environments
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% ========================================
% Document Information
% ========================================
\newcommand{\dissertationtitle}{Evolution of Cooperation\\under Environmental Variability}
\newcommand{\authorname}{Masaaki Inaba}
\newcommand{\submissiondate}{March 2026}
\newcommand{\universityname}{University of Tsukuba}
\newcommand{\academicgroup}{Graduate School of Science and Technology}
\newcommand{\degreeprogram}{Degree Programs in Systems and Information Engineering}

% ========================================
% Document Body
% ========================================
\begin{document}

% ========================================
% Cover Page
% ========================================
\begin{titlepage}
    \centering
    \vspace*{5cm}
    {\fontsize{24}{26}\selectfont\bfseries \dissertationtitle\par}
    \vspace*{\fill}
    {\fontsize{20}{22}\selectfont \authorname\par}
    \vspace{1cm}
    {\fontsize{20}{22}\selectfont \submissiondate\par}
    \vspace*{3cm}
\end{titlepage}

% ========================================
% Inside Cover
% ========================================
\newpage
\thispagestyle{empty}
\begin{center}
    \vspace*{5cm}
    {\fontsize{24}{26}\selectfont\bfseries \dissertationtitle\par}
    \vspace*{\fill}
    {\fontsize{20}{22}\selectfont \academicgroup\par}
    {\fontsize{20}{22}\selectfont \degreeprogram\par}
    {\fontsize{20}{22}\selectfont \universityname\par}
    \vspace*{\fill}
    {\fontsize{20}{22}\selectfont \authorname\par}
    \vspace{1cm}
    {\fontsize{20}{22}\selectfont \submissiondate\par}
    \vspace*{3cm}
\end{center}
\newpage

% ========================================
% Abstract
% ========================================
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

\begin{CJK}{UTF8}{min}
{\color{red}
博論全体の概要を記載する予定です。
現時点 (プレ予備審査時点) では Chapter 2 と 4 の Abstract を併記しています。
}
\end{CJK}
\vspace*{1cm}

\begin{CJK}{UTF8}{min}
{\color{red}
Chapter 2 の Abstract
}
\end{CJK}

The evolutionary process that led to the emergence of modern human behaviors during the Middle Stone Age in Africa remains enigmatic.
While various hypotheses have been proposed, we offer a new perspective that integrates the variability selection hypothesis (VSH) with the evolution of cooperation among human groups.
The VSH suggests that human adaptability to fluctuating environments was a primary force driving the development of key evolutionary traits.
However, the mechanisms by which environmental variability (EV) influenced human evolution, particularly the emergence of large-scale and complex cooperative behaviors, are not yet fully understood.
To explore the connection between intensified EV and the evolution of intergroup cooperation, we analyzed three stochastic models of EV:
(i) Regional Variability (RV), where resource-rich areas shift while overall resource levels remain stable;
(ii) Universal Variability (UV), where overall resource levels fluctuate but resource-rich areas remain stable; and
(iii) Combined Variability (CV), where both resource-rich areas shift and overall resource levels fluctuate.
Our results show that RV strongly promotes cooperation, while UV has a comparatively weaker effect.
Additionally, our findings indicate that the coevolution of cooperation and network structures is crucial for EVs to effectively promote cooperation.
This study proposes a novel causal link between EV and the evolution of cooperation, potentially setting a new direction for theoretical and empirical research in this field.

\vspace*{1cm}
\begin{CJK}{UTF8}{min}
{\color{red}
Chapter 4 の Abstract
}
\end{CJK}

Cooperation is fundamental to human societies.
While several basic theoretical mechanisms underlying its evolution have been established, research addressing more realistic settings remains underdeveloped.
Drawing on the hypothesis that intensified environmental fluctuations influenced early behavioral evolution in humans during the Middle Stone Age in Africa, we examine the effects of environmental variability and human mobility on the evolution of cooperation.
In our model, the variability is represented by randomly moving resource-rich spots across a two-dimensional space, and the mobility is represented by resource-seeking migration of agents.
The agents interact cooperatively or competitively for resources while adopting their behavioral strategies from more successful neighbors.
Through extensive simulations of this model, we reveal three key findings:
(i) with sufficient agent mobility, even modest environmental variability promotes cooperation, but further variability does not enhance cooperation;
(ii) with any level of environmental variability, agent mobility promotes cooperation; and
(iii) these effects occur because the joint effect of environmental variability and agent mobility disrupts defector groups in resource-rich areas, forming cooperator groups at those sites.
Although previous studies examined environmental variability and mobility separately, to the best of our knowledge, this is the first study to analyze their joint effects on the evolution of cooperation.
These findings suggest that environmental variability can promote cooperative group formation without enhanced cognitive abilities, providing new insights into the evolution of human cooperation and, by extension, sociality.

% ========================================
% Table of contents
% ========================================
\tableofcontents
\listoffigures
\listoftables

% ========================================
% Main Content
% ========================================
\include{chapters/1_introduction}
\include{chapters/2_base_model}
\include{chapters/3_2Lvl_model}
\include{chapters/4_2D_model}
\include{chapters/5_conclusion}

% ========================================
% Appendix
% ========================================
\addcontentsline{toc}{chapter}{Appendix}
\chapter*{Appendix}

\section*{Data and code availability}
All data, as well as the code required to run the simulations and generate the figures, are available at

\begin{itemize}
    \item Chapter 2: https://github.com/mas178/Inaba2024
    \item Chapter 3: https://github.com/mas178/Inaba2025-2Lvl
    \item Chapter 4: https://github.com/mas178/Inaba2025-2D
\end{itemize}

\section*{Supplementary material for Chapter 3}

\section*{Supplementary material for Chapter 4}

% ========================================
% Bibliography
% ========================================
\addcontentsline{toc}{chapter}{Bibliography}
\printbibliography[heading=bibintoc]

% ========================================
% Acknowledgments
% ========================================
\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}


\end{document}