\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

I would like to express my gratitude to those who contributed to the completion of this dissertation.

First and foremost, I wish to thank my primary supervisor, Professor Eizo Akiyama, for his guidance throughout my five years of doctoral studies.
During the formative stages of my graduate education, he provided a necessary foundation in the principles and methodology of scholarly research.
I am especially indebted to him for the consistent engagement with my research; he remained the first to understand my ideas, at times a rigorous critic, and always a dedicated supporter.
I also acknowledge his contributions as a co-author of the two published articles that form the basis of this dissertation.

I am also grateful to my secondary supervisors, Professor Ryoji Sawa and Professor Tuan Phung-Duc, for their critical questions and constructive advice.
Their feedback provided an invaluable opportunity to further deepen the discussion and insights of this study.

Furthermore, I would like to thank the members of the examination committee for their insightful comments.
I am grateful to Assistant Professor Hidekazu Anno for providing critical feedback from the perspective of an expert outside the field of study, which helped situate the research within a broader academic context.
I also thank Professor Hitoshi Yamamoto, whose extensive knowledge and experience in the evolution of cooperation proved indispensable in refining the theoretical rigor and academic significance of this work.

Finally, I acknowledge the institutional and academic support provided by the University of Tsukuba, which made this research possible.
